An individual differences perspective on the development of pragmatic abilities in the preschool years

### 2022-04-24 paper/manuscript.Rmd
add preprint link

abstract: |
  Pragmatic abilities are fundamental to successful language use and learning. Individual differences studies contribute to understanding the psychological processes involved in pragmatic reasoning. Small sample sizes, insufficient measurement tools, and a lack of theoretical precision have hindered progress, however. Three studies addressed these challenges in a sample of three- to five-year-old German-speaking children (N = 228, 121 female). Studies 1 and 2 assessed the psychometric properties of six pragmatics tasks. Study 3 investigated relations among pragmatics tasks and between pragmatics and other cognitive abilities. The shared cognitive processes were formalized in a computational cognitive model of pragmatic reasoning. The tasks were found to measure stable variation between individuals on a shared latent construct relating to pragmatic inference.

keywords          : "Pragmatics, language development, individual differences, cognitive modeling"
wordcount         : "8491"

@@

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_docx
---

@@

Communication predates language. Before children produce their first words, they communicate with the world around them using vocalizations and gestures [@bruner1974communication; @bates1979emergence]. The process of language learning recruits many of the social-cognitive processes that underlie pre-verbal communication [@bohn2019pervasive; @tomasello2009constructing; @clark2009first]. Even for proficient language users, communication not reducible to the words being exchanged. The common thread running through these different aspects of human communication is its inferential nature: what a speaker means -- verbal or otherwise -- is underdetermined by the parts that make up the utterance. It takes contextual social inferences to recover the intended meaning [@grice1991studies; @clark1996using; @sperber2001relevance; @levinson2000presumptive; @bohn2018common]. These inferences are often referred to as *pragmatic* inferences.

The development of pragmatics has been widely studied in recent years [for a recent review see @bohn2019pervasive]. This research covers a range of different phenomena ranging from so-called *pure pragmatics* [@matthews2014pragmatic] in non-verbal communication in infancy to sophisticated linguistic inferences developing much later [@huang2009online; @papafragou2016scalar]. A growing portion of this work is devoted to studying individual differences [@matthews2018individual; @wilson2021pragmatic]. The motivation behind the move to study individual variation is twofold: on the one hand, individual differences offer insights into the underlying psychological processes. If two phenomena (e.g. pragmatic reasoning and executive functions) vary together this is consistent with shared cognitive processes [@kidd2018individual; @matthews2018individual; @wilson2022novel], though it is not definitive evidence for such a claim. On the other hand, deficits in pragmatic abilities have been linked to behavioral problems and forms of language impairment [@helland2014stable].

In their recent review, @matthews2018individual identified three issues that significantly limit what we can learn from individual differences research on pragmatic abilities. First, most studies have insufficient sample sizes so that small and medium sized correlations among pragmatics tasks and between pragmatics tasks and measures for other cognitive abilities cannot be reliably detected [mirroring issues in estimating correlations across other fields, @schonbrodt2013sample]. Second, the tasks used to assess pragmatic abilities often have poor or unknown psychometric properties. For example, many tasks only have a single trial and are therefore unable to capture variation between children [see also @enkavi2019large]. Reliability is not assessed, making it unclear if the task captures stable characteristics [@russell2008assessing; @flake2020measurement]. Third, the cognitive processes underlying pragmatic inferences in a particular task are underspecified. As a consequence, there is often no clear rationale for why a particular target task should correlate with another cognitive measure.

In search of a better understanding of individual variation in pragmatic ability, the studies presented here directly address these issues. We identified six pragmatic reasoning tasks in children between three and five years of age and investigated their psychometric properties, in particular their re-test reliability. Reliable tasks are a necessary precondition for meaningful individual differences research [@hedge2018reliability; @fried2018measurement]. Next, we investigated the relations among different pragmatic reasoning tasks as well as between pragmatic reasoning and other cognitive abilities in a sample large enough to detect small to medium sized correlations.

 Finally, we introduced computational cognitive models of pragmatic reasoning to the study of individual differences. Computational cognitive models formalize hypotheses about cognitive processes that could underlie pragmatic reasoning; thus, the use of these models provides a substantive theoretical account for why certain pragmatic reasoning tasks should be related to one another. Here, we use the formalism introduced by the Rational Speech Act (RSA) framework [@frank2012predicting; @goodman2016pragmatic]. RSA models see pragmatic inferences as a special case of (Bayesian) social reasoning. A pragmatic listener interprets an utterance by assuming it was produced by a cooperative speaker. The speaker tries to be informative, that is, they provide messages that would increase the probability that the listener will recover their intended meaning. The informativeness of an utterance arises from a contrastive inference in which the effects of multiple – plausible – utterances are compared. We assume that this inference process is shared by some of the pragmatics tasks involved in this study and can thus be used to account for individual differences (see below).

The six tasks we selected were developmental adaptations of referential communication games inspired by research in experimental pragmatics [@noveck2008experimental; @noveck2004experimental]. They all share a common trial-by-trial structure in which the test event always involved an agent producing an ambiguous utterance that the child had to resolve using pragmatic reasoning. This structure allowed us to run multiple trials per task, increasing reliability. We grouped the tasks into two broad categories (Figure \@ref(fig:fig1)). On the one hand, *utterance-based tasks* asked children to derive inferences from the words and gestures the speaker produced in context. On the other hand, *common ground/discourse-based tasks* asked children to derive inferences from the social interaction that preceded the utterance.

For the utterance-based category, we selected mutual exclusivity, informativeness inference, and ad-hoc implicature tasks. "Mutual exclusivity"^[Following @lewis2020role, we use the term "mutual exclusivity" as a convenient term to denote a specific task. This term is also related to a particular theoretical account of the phenomenon [@markman1990constraints], but we do not presuppose that specific account.] describes the phenomenon that children tend to map a novel word to an unknown object [@markman1988children; @halberda2003development; @bion2013fast; @merriman1989mutual; @clark1988logic; @lewis2020role]. Informativeness inferences describe situations in which children identify the referent of a novel word by assuming that the speaker is trying to be informative. Being informative translates to using words that have smaller extensions in context [@frank2014inferring]. Ad-hoc implicature describes inferences that ask the child to contrast an utterance with alternatives that the speaker could have used but did not [@stiller2015ad; @yoon2019role; @katsos2011pragmatic].

@@

After finding in Study 1 and 2 that most of the pragmatics tasks showed relatively good re-test reliability, we studied a larger sample of children in Study 3 to investigate relations between the three utterance-based tasks. We focused on these tasks for theoretical reasons: as noted above, we assume that -- computationally -- they share a common contrastive inference process. We formalize these assumptions in a computational cognitive model which we then use to study individual differences in this alleged process. Study 3 also included tasks for executive functions and analogical reasoning. Across analytical approaches, we found systematic relations among the pragmatics tasks as well as between pragmatics and executive functions, but not analogical reasoning. In the discussion, we use the structure of the cognitive model to speculate about the psychological processes shared between pragmatics and executive functions.

@@

Study 1 focused on the psychometric properties of four pragmatics tasks, in particular, their re-test reliability. Two of the tasks were from the utterance-based group and two from the common ground/discourse-based group. As a fifth task, we included a measure of executive functions. Methods and sample size were pre-registered at https://osf.io/6a723. All analysis scripts and data files can be found in the following repository: https://github.com/manuelbohn/pragBat. The same repository also contains the code to run the experiments.

@@

To estimate developmental effects in each task, we fit a GLMM predicting correct responses (0/1) by age (in years, centered at the mean) and trial number (also centered). The model included random intercepts for each participant and random slopes for trial within participants (model notation in `R`: `correct ~ age + trial + (trial|id)` )^[We pre-registered the inclusion of random intercepts for item. We deviate from this here because the order of items was fixed and the same for all participants so that trial and item were confounded.]. For each task, we inspected and visualized the posterior distribution (mean and 95% Credible Interval (CrI)) for the age estimate.

We assessed re-test reliability in two ways. First, for each task we computed the proportion of correct trials for each individual in the two test sessions and then used Pearson correlations to quantify re-test reliability. Second, we used a GLMM based approach suggested by @rouder2019psychometrics. Here, a GLMM was fitted to the trial-by-trial data for each task with a fixed effect of age (in years, centered at the mean), a random intercept for each participant and a random slope for test day (`correct ~ age + (0+test_day|id)`)^[The notation `0+test_day` yields a separate intercept estimate for each test day and subject instead of an intercept estimate for day 1 and a slope for the difference between day 1 and day 2. As a consequence, the model estimates the correlation between the two test days instead of a correlation between an intercept and the slope for test day.]. The model yields a participant specific estimate for each test day and also estimates the correlation between the two. This correlation can be interpreted as the re-test reliability. This approach has several advantages. First, it uses trial-by-trial data and avoids information loss that comes with data aggregation. Second, it uses hierarchical shrinkage to obtain better participant-specific estimates. Finally, it allows us to get an age-independent estimate for reliability. One worry when assessing re-test reliability in developmental studies is that re-test correlations can be high because of domain general cognitive gains and not because of task-specific individual differences. By including age as a fixed effect in the model, the estimates for each participant are independent of age and so is the correlation between estimates for the two test days -- the re-test reliability.

@@

Based on these results of Study 1, we compiled a new set of tasks for Study 2. We retained the mutual exclusivity and card sorting tasks because of the interesting relation between the two found in Study 2. We simplified the informativeness inference task to be more age appropriate with the hope of inducing more variation in performance. We removed the speaker preference and discourse novelty tasks -- despite their excellent re-test reliability -- because they seemed to be unrelated to one another and also unrelated to the other tasks. We also added new tasks focused on ad-hoc implicature and discourse continuity. As noted in the introduction, we had theoretical reasons to expect the ad-hoc implicature task to be related to the mutual exclusivity and informativeness inference tasks. We had no such strong predictions for the discourse continuity task.

@@

The reliability estimates from Study 1 and 2 helped us plan the sample size for Study 3. The focal tasks had a re-test reliability around 0.7. Because the highest plausible correlation between two tasks is the product of their reliabilities (higher correlations would mean that the task is more strongly related to a different task than to itself), the highest we could expect were correlations between two tasks around 0.7 * 0.7 = 0.49. We planned our sample so that we could detect correlations between two tasks of 0.3 with 95% power^[The first author drafted a pre-registration and shared it with the last author but forgot to register it at OSF. Thus, the study was not officially pre-registered.]. Data, analysis scripts and experiment code can be found in the associated online repository.

@@

$$P_{L_1}(r|u)\propto P_{S_1}(u|r) \cdot P(r)$$
In the above equation, the listener ($P_{L_1}$) is trying to infer the speaker’s ($P_{S_1}$) intended referent r by imagining what a rational speaker would say, given the referent they are trying to communicate and combining it with the listener’s prior expectations about the referent $P(r)$^[We assumed a uniform prior distribution over potential referents.]. The speaker is an approximately rational Bayesian actor (with degree of rationality $\alpha$) who produces utterances as a function of their informativity.

>>>>

abstract: |
  Pragmatic abilities are fundamental to successful language use and learning. Individual differences studies contribute to understanding the psychological processes involved in pragmatic reasoning. Small sample sizes, insufficient measurement tools, and a lack of theoretical precision have hindered progress, however. Three studies addressed these challenges in three- to five-year-old German-speaking children (N = 228, 121 female). Studies 1 and 2 assessed the psychometric properties of six pragmatics tasks. Study 3 investigated relations among pragmatics tasks and between pragmatics and other cognitive abilities. The tasks were found to measure stable variation between individuals. Via a computational cognitive model, individual differences were traced back to a latent pragmatics construct. This presents the basis for understanding the relations between pragmatics and other cognitive abilities.

keywords          : "Pragmatics, language development, individual differences, cognitive modeling"
wordcount         : "8558"

@@

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

@@

Communication predates language. Before children produce their first words, they communicate with the world around them using vocalizations and gestures [@bruner1974communication; @bates1979emergence]. The process of language learning recruits many of the social-cognitive processes that underlie pre-verbal communication [@bohn2019pervasive; @tomasello2009constructing; @clark2009first]. Even for proficient language users, communication is not reducible to the words being exchanged. The common thread running through the different aspects of human communication is its inferential nature: what a speaker means -- verbal or otherwise -- is underdetermined by the parts that make up the utterance. It takes contextual social inferences, often referred to as *pragmatic* inferences, to recover the intended meaning [@grice1991studies; @clark1996using; @sperber2001relevance; @levinson2000presumptive; @bohn2018common].

The development of pragmatics has been widely studied in recent years [for a recent review see @bohn2019pervasive]. This research covers a range of different phenomena ranging from so-called *pure pragmatics* [@matthews2014pragmatic] in non-verbal communication in infancy to sophisticated linguistic inferences developing much later [@huang2009online; @papafragou2016scalar]. A growing portion of this work is devoted to studying individual differences [@matthews2018individual; @wilson2021pragmatic]. The motivation behind the move to study individual variation is twofold: first, individual differences offer insights into the underlying psychological processes. If two phenomena (e.g. pragmatic reasoning and executive functions) vary together this is consistent with shared cognitive processes [@kidd2018individual; @matthews2018individual; @wilson2022novel], though it is not definitive evidence for such a claim. Second, deficits in pragmatic abilities have been linked to maladaptive behavioral patterns and forms of language impairment [@helland2014stable].

In their recent review, @matthews2018individual identified three issues that significantly limit what we can learn from individual differences research on pragmatic abilities. First, most studies have insufficient sample sizes so that small and medium sized correlations among pragmatics tasks and between pragmatics tasks and measures for other cognitive abilities cannot be reliably detected [mirroring issues in estimating correlations across other fields, @schonbrodt2013sample]. Second, the tasks used to assess pragmatic abilities often have poor or unknown psychometric properties. For example, many tasks only have a single trial and are therefore unable to capture variation between children [see also @enkavi2019large]. Furthermore, reliability is not assessed, making it unclear if the task captures stable characteristics [@russell2008assessing; @flake2020measurement]. Third, the cognitive processes underlying pragmatic inferences in a particular task are underspecified. As a consequence, there is often no clear rationale for why a particular target task should correlate with another cognitive measure.

In search of a better understanding of individual variation in pragmatic ability, the studies presented here directly address these issues. We identified six pragmatic reasoning tasks in children between three and five years of age and investigated their psychometric properties, in particular their re-test reliability. Reliable tasks are a necessary precondition for meaningful individual differences research [@hedge2018reliability; @fried2018measurement]. Next, we investigated the relations among different pragmatic reasoning tasks as well as between pragmatic reasoning and other cognitive abilities in a sample large enough to detect small to medium sized correlations. For this purpose, we introduced computational cognitive models of pragmatic reasoning to the study of individual differences. Computational cognitive models formalize hypotheses about cognitive processes that could underlie pragmatic reasoning; thus, the use of these models provides a substantive theoretical account of why certain pragmatic reasoning tasks should be related to one another. Here, we use the formalism introduced by the Rational Speech Act (RSA) framework [@frank2012predicting; @goodman2016pragmatic]. RSA models see pragmatic inferences as a special case of (Bayesian) social reasoning. A pragmatic listener interprets an utterance by assuming it was produced by a cooperative speaker. The speaker tries to be informative, that is, they provide messages that would increase the probability that the listener will recover their intended meaning. The informativeness of an utterance arises from a contrastive inference in which the effects of multiple – plausible – utterances are compared. We assume that this inference process is shared by some of the pragmatics tasks involved in this study and can thus be used to account for individual differences (see below).

The six tasks we selected were developmental adaptations of referential communication games inspired by research in experimental pragmatics [@noveck2008experimental; @noveck2004experimental]. They all share a common trial-by-trial structure in which the test event always involved an agent producing an ambiguous utterance that the child had to resolve using pragmatic reasoning. This structure allowed us to run multiple trials per task, increasing reliability. We grouped the tasks into two broad categories (Figure \@ref(fig:fig1)). *Utterance-based tasks* asked children to derive inferences from the words and gestures the speaker produced in context. *Common ground/discourse-based tasks* asked children to derive inferences from the social interaction that preceded the utterance.

For the utterance-based category, we selected mutual exclusivity, informativeness inference, and ad-hoc implicature tasks. "Mutual exclusivity" describes the phenomenon that children tend to map a novel word to an unknown object [@markman1988children; @halberda2003development; @bion2013fast; @merriman1989mutual; @clark1988logic; @lewis2020role]. Following @lewis2020role, we use the term "mutual exclusivity" as a convenient term to denote a specific task. This term is also related to a particular theoretical account of the phenomenon [@markman1990constraints], but we do not presuppose that specific account. Informativeness inferences describe situations in which children identify the referent of a novel word by assuming that the speaker is trying to be informative. Being informative translates to using words that reduce ambiguity and help the listener to recover the intended meaning [@frank2014inferring]. Ad-hoc implicature describes inferences that ask the child to contrast an utterance with alternatives that the speaker could have used but did not [@stiller2015ad; @yoon2019role; @katsos2011pragmatic].

@@

Study 1 and 2 explored the re-test reliability of the pragmatics tasks and found it to be  relatively good. Study 3 tested a larger sample of children to investigate relations between the three utterance-based tasks. We focused on these tasks for theoretical reasons: as noted above, we assume that -- computationally -- they share a common contrastive inference process. We formalize these assumptions in a computational cognitive model which we then use to study individual differences in this alleged process. Study 3 also included tasks for executive functions and analogical reasoning. Across analytical approaches, we found systematic relations among the pragmatics tasks as well as between pragmatics and executive functions, but not analogical reasoning. In the discussion, we use the structure of the cognitive model to speculate about the psychological processes shared between pragmatics and executive functions.

Taken together, this study introduces a set of tasks that reliably measure individual differences in pragmatic abilities in the preschool years. In addition, it introduces a new (formal) theoretical framework that help us understand individual differences on a process level and, with that, suggests answers to why pragmatic abilities relate to other cognitive abilities.

@@

Study 1 focused on the psychometric properties of four pragmatics tasks, in particular, their re-test reliability. We chose our sample size so that we would detect medium to high re-test correlations with sufficient power. Two of the tasks were from the utterance-based group and two from the common ground/discourse-based group. This design allowed us to explore whether tasks within one group are more related to one another than between groups. As a fifth task, we included a measure of executive functions. Methods and sample size were pre-registered at https://osf.io/6a723. All analysis scripts and data files can be found in the following repository: https://github.com/manuelbohn/pragBat. The same repository also contains the code to run the experiments.

@@

To estimate developmental effects in each task, we fit a GLMM predicting correct responses (0/1) by age (in years, centered at the mean) and trial number (also centered). The model included random intercepts for each participant and random slopes for trial within participants (model notation in `R`: `correct ~ age + trial + (trial|id)` ). We pre-registered the inclusion of random intercepts for item. We deviate from this here because the order of items was fixed and the same for all participants so that trial and item were confounded for each task. For each task, we inspected and visualized the posterior distribution (mean and 95% Credible Interval (CrI)) for the age estimate.

We assessed re-test reliability in two ways. First, for each task we computed the proportion of correct trials for each individual in the two test sessions and then used Pearson correlations to quantify re-test reliability. Second, we used a GLMM based approach suggested by @rouder2019psychometrics. Here, a GLMM was fitted to the trial-by-trial data for each task with a fixed effect of age (in years, centered at the mean), a random intercept for each participant and a random slope for test day (`correct ~ age + (0+test_day|id)`). The notation `0+test_day` yields a separate intercept estimate for each test day and subject instead of an intercept estimate for day 1 and a slope for the difference between day 1 and day 2. As a consequence, the model estimates the correlation between the two test days instead of a correlation between an intercept and the slope for test day. The correlation between test days can be interpreted as the re-test reliability. This approach has several advantages. First, it uses trial-by-trial data and avoids information loss that comes with data aggregation. Second, it uses hierarchical shrinkage to obtain better participant-specific estimates. Finally, it allows us to get an age-independent estimate for reliability. One worry when assessing re-test reliability in developmental studies is that re-test correlations can be high because of domain general cognitive gains and not because of task-specific individual differences. By including age as a fixed effect in the model, the estimates for each participant are independent of age and so is the correlation between estimates for the two test days -- the re-test reliability.

@@

The goal of Study 2 was to assess the re-test reliability in a new set of tasks. We retained the mutual exclusivity and card sorting tasks because of the interesting relation between the two found in Study 2. We simplified the informativeness inference task to be more age appropriate with the hope of inducing more variation in performance. We removed the speaker preference and discourse novelty tasks -- despite their excellent re-test reliability -- because they seemed to be unrelated to one another and also unrelated to the other tasks. We also added new tasks focused on ad-hoc implicature and discourse continuity. As noted in the introduction, we had theoretical reasons to expect the ad-hoc implicature task to be related to the mutual exclusivity and informativeness inference tasks. We had no such strong predictions for the discourse continuity task.

@@

The reliability estimates from Study 1 and 2 helped us plan the sample size for Study 3. The focal tasks had a re-test reliability around 0.7. Because the highest plausible correlation between two tasks is the product of their reliabilities (higher correlations would mean that the task is more strongly related to a different task than to itself), the highest we could expect were correlations between two tasks around 0.7 * 0.7 = 0.49. We planned our sample so that we could detect correlations between two tasks of 0.3 with 95% power. The first author drafted a pre-registration and shared it with the last author but forgot to register it at OSF. Thus, the study was not officially pre-registered. Data, analysis scripts and experiment code can be found in the associated online repository.

@@

$$P_{L_1}(r|u)\propto P_{S_1}(u|r) \cdot P(r)$$
In the above equation, the listener ($P_{L_1}$) is trying to infer the speaker’s ($P_{S_1}$) intended referent $r$ by imagining what a rational speaker would say, given the referent they are trying to communicate and the listener’s prior expectations about the referent $P(r)$ (which we assumed to be uniform over potential referents). The speaker is an approximately rational Bayesian actor (with degree of rationality $\alpha$) who produces utterances as a function of their informativity.
